Master Thesis

### 2013-06-13 the-last-part.tex
Switched to the report class, start using theorem envs

\section{The Last Part}

In this section, we will attempt to incorporate accounts of several
linguistic phenomena in a single framework, highlight the modularity
issues this brings up and propose some solutions to them.

\subsection{Negation}

We will start with an account of the paired grammatical words mechanism
for negation in French. Recall the EPTDs we have seen
in~\ref{sssec:frigram}, repeated here on Figure~\ref{fig:ig-neg-rep}.

@@

If we suppose the presence of the appropriate lexical items typed as in
the examples of~\ref{ssec:acg}, we have a type system in which we can
type the syntactic terms corresponding to the sentences
(\ref{ex:good-neg-double}), (\ref{ex:good-neg-embed}) and
(\ref{ex:good-neg-rel}) while making it impossible to type the
structures corresponding to sentences (\ref{ex:bad-neg-noneg}) and
({\ref{ex:bad-neg-rel}}).

@@

\subsection{Extraction}
\label{ssec:extraction}

@@

\subsection{Agreement}

@@

In the next subsection, we will see what it takes to enforce all of the
constraints above (negation, extraction and agreement) at the same time,
which will give us some idea on how to go about building a grammar which
handles a wide variety of such phenomena.

\subsection{Putting It All Together}

@@

As of right now, we have introduced this kind of rule only as
meta-notation. This means that the real underlying grammar that will end
up being used by some algorithm still suffers from the same exponential
cardinality of the signature. Fortunately, there exists a well-studied
construction in type theory that lets us assign to terms generic types
such as the one defined in the template above. This construction is the
\emph{dependent product} and its utility in ACGs has already been
noted.\footnote{See \cite{de2007two} for its introduction into the
  domain of ACGs, \cite{de2007type} for a small example grammar, which
  handles the interaction between number and gender agreement and
  coordination, and finally see \cite{pogodalla2012controlling} for
  using dependent types to handle several constraints regarding
  extraction (including the one we covered in~\ref{ssec:extraction}).}

While extending the type system by including dependent types is
definitely useful and it helps to solve the problem of having our
signatures grow exponentially in size, it does not solve our problem
with the complexity of the new grammar. Granted, the new meta-type
written in the dependent style is more readable and easier to reason
about than the enumeration of 12 basic types we presented at the
beginning of this subsection, however, the result is still complex in
that it forces us to describe the behavior of three independent
mechanisms in one place. This one place happens to be a wordform which
has seemingly little connection to these mechanisms, the preposition
$de$. We can therefore imagine that it will participate in many other
phenomena and that the type we assign to it will grow beyond our
abilities to reason about it effectively.

@@

\subsection{Intersections of Types}

@@

\subsection{Intersections of Languages}

@@

The definition of ACGs given above is quite practical in that it is
complete, it gives a precise account of what set of languages are
generated by ACGs (abstract and object languages for which there exist
the quadruples described above). This contrasts with the presentation of
ACGs that we gave in~\ref{ssec:acg}, where we defined ACGs as
collections of signatures with associated distinguished types and
connected with lexicons. Our definition only introduced the requisite
machinery, but delayed the delimitation of the way languages can be
defined. Leaving this question open will let us answer it now by
providing an extension/generalization of ACGs which will help us
overcome the modularity issues that have been the theme of this
section.\footnote{Incidentally, this was not the reason for the way we
  presented ACGs in~\ref{ssec:acg}. We chose this style as it grouped
  together the information in a more natural way. We approach signatures
  as descriptions of domains (strings, syntax, semantics), each one
  being associated with a distinguished type telling us which objects
  the domain has set out to describe (strings, sentences, truth
  values). Having distinguished types directly attached to signatures
  allows us to omit them from the ACG quadruples. However, this only
  leaves us with triples containing the two signatures and the lexicon
  connecting them, and since we can infer the signatures themselves from
  the domain and the range of the lexicon, we do not have much
  motivation to introduce these tuples. Instead of composing ACGs, we
  compose just the lexicons.

@@

\subsubsection{Graphical Abstract Categorial Grammars}

@@

\subsubsubsection{Formal Properties of Graphical ACGs}

Observation: The set of intrinsic languages definable by graphical ACGs
is the same as the set of abstract languages definable by ACGs.

Theorem: The set of extrinsic languages definable by graphical ACGs is
the set of object languages definable by ACGs closed under intersection
and transformation by a lexicon.

Proof:

First, we prove that an extrinsic language can be always constructed
from object languages by intersections and transformations by
lexicons. Let us consider any graphical ACG $\mathcal{G} = \mathopen{<}
G, \Sigma, S, \mathcal{L} \mathclose{>}$ and any topological ordering
$v_1, \ldots, v_n$ of the nodes of $G$ (an ordering such that more
abstract nodes always precede less abstract ones). Before we start, we
will just note that the all the intrinsic languages defined by $v_i$ are
abstract languages which are a special case of object languages.

@@

  \begin{align*}
    \mathcal{E}_{\mathcal{G}'}(v) &= \mathcal{I}_{\mathcal{G}'}(v) \cap
    \mathcal{L}(\mathcal{E}_{\mathcal{G}'}(u)) \\
    &= \mathcal{L}(\mathcal{E}_{\mathcal{G}'}(u)) \\
    &= \mathcal{L}(\mathcal{E}_{\mathcal{G}}(u)) \\
    &= \mathcal{L}(L)
  \end{align*}

\end{itemize}

Corollary: The set of extrinsic languages definable by graphical ACGs is
the same as the set of object languages definable by ACGs if and only if
the set of object languages definable by ACGs is closed under
intersection.

>>>>

\chapter{The Last Part}

In this chapter, we will attempt to incorporate accounts of several
linguistic phenomena in a single framework, highlight the modularity
issues this brings up and propose some solutions to them.

\section{Negation}

We will start with an account of the paired grammatical words mechanism
for negation in French. Recall the EPTDs we have seen
in~\ref{ssec:frigram}, repeated here on Figure~\ref{fig:ig-neg-rep}.

@@

If we suppose the presence of the appropriate lexical items typed as in
the examples of~\ref{sec:acg}, we have a type system in which we can
type the syntactic terms corresponding to the sentences
(\ref{ex:good-neg-double}), (\ref{ex:good-neg-embed}) and
(\ref{ex:good-neg-rel}) while making it impossible to type the
structures corresponding to sentences (\ref{ex:bad-neg-noneg}) and
({\ref{ex:bad-neg-rel}}).

@@

\section{Extraction}
\label{sec:extraction}

@@

\section{Agreement}

@@

In the next section, we will see what it takes to enforce all of the
constraints above (negation, extraction and agreement) at the same time,
which will give us some idea on how to go about building a grammar which
handles a wide variety of such phenomena.

\section{Putting It All Together}

@@

As of right now, we have introduced this kind of rule only as
meta-notation. This means that the real underlying grammar that will end
up being used by some algorithm still suffers from the same exponential
cardinality of the signature. Fortunately, there exists a well-studied
construction in type theory that lets us assign to terms generic types
such as the one defined in the template above. This construction is the
\emph{dependent product} and its utility in ACGs has already been
noted.\footnote{See \cite{de2007two} for its introduction into the
  domain of ACGs, \cite{de2007type} for a small example grammar, which
  handles the interaction between number and gender agreement and
  coordination, and finally see \cite{pogodalla2012controlling} for
  using dependent types to handle several constraints regarding
  extraction (including the one we covered in~\ref{sec:extraction}).}

While extending the type system by including dependent types is
definitely useful and it helps to solve the problem of having our
signatures grow exponentially in size, it does not solve our problem
with the complexity of the new grammar. Granted, the new meta-type
written in the dependent style is more readable and easier to reason
about than the enumeration of 12 basic types we presented at the
beginning of this section, however, the result is still complex in
that it forces us to describe the behavior of three independent
mechanisms in one place. This one place happens to be a wordform which
has seemingly little connection to these mechanisms, the preposition
$de$. We can therefore imagine that it will participate in many other
phenomena and that the type we assign to it will grow beyond our
abilities to reason about it effectively.

@@

\section{Intersections of Types}

@@

\section{Intersections of Languages}

@@

The definition of ACGs given above is quite practical in that it is
complete, it gives a precise account of what set of languages are
generated by ACGs (abstract and object languages for which there exist
the quadruples described above). This contrasts with the presentation of
ACGs that we gave in~\ref{sec:acg}, where we defined ACGs as
collections of signatures with associated distinguished types and
connected with lexicons. Our definition only introduced the requisite
machinery, but delayed the delimitation of the way languages can be
defined. Leaving this question open will let us answer it now by
providing an extension/generalization of ACGs which will help us
overcome the modularity issues that have been the theme of this
chapter.\footnote{Incidentally, this was not the reason for the way we
  presented ACGs in~\ref{sec:acg}. We chose this style as it grouped
  together the information in a more natural way. We approach signatures
  as descriptions of domains (strings, syntax, semantics), each one
  being associated with a distinguished type telling us which objects
  the domain has set out to describe (strings, sentences, truth
  values). Having distinguished types directly attached to signatures
  allows us to omit them from the ACG quadruples. However, this only
  leaves us with triples containing the two signatures and the lexicon
  connecting them, and since we can infer the signatures themselves from
  the domain and the range of the lexicon, we do not have much
  motivation to introduce these tuples. Instead of composing ACGs, we
  compose just the lexicons.

@@

\subsection{Graphical Abstract Categorial Grammars}

@@

\subsubsection{On the Expressivity of Graphical ACGs}

\begin{observation}
  The set of intrinsic languages definable by graphical ACGs is the same
  as the set of abstract languages definable by ACGs.
\end{observation}

\begin{theorem}
  The set of extrinsic languages definable by graphical ACGs is the set
  of object languages definable by ACGs closed under intersection and
  transformation by a lexicon.
\end{theorem}

\begin{proof}
  First, we prove that an extrinsic language can be always constructed
  from object languages by intersections and transformations by
  lexicons. Let us consider any graphical ACG $\mathcal{G} =
  \mathopen{<} G, \Sigma, S, \mathcal{L} \mathclose{>}$ and any
  topological ordering $v_1, \ldots, v_n$ of the nodes of $G$ (an
  ordering such that more abstract nodes always precede less abstract
  ones). Before we start, we will just note that the all the intrinsic
  languages defined by $v_i$ are abstract languages which are a special
  case of object languages.

@@

    \begin{align*}
      \mathcal{E}_{\mathcal{G}'}(v) &= \mathcal{I}_{\mathcal{G}'}(v) \cap
      \mathcal{L}(\mathcal{E}_{\mathcal{G}'}(u)) \\
      &= \mathcal{L}(\mathcal{E}_{\mathcal{G}'}(u)) \\
      &= \mathcal{L}(\mathcal{E}_{\mathcal{G}}(u)) \\
      &= \mathcal{L}(L)
    \end{align*}
  \end{itemize}
\end{proof}

\begin{corollary}
  The set of extrinsic languages definable by graphical ACGs is the same
  as the set of object languages definable by ACGs if and only if the
  set of object languages definable by ACGs is closed under
  intersection.
\end{corollary}
